---
title: SALMON Bulk RNA Differential expresion analysis with DESeq2 & edgeR
date: "Last Updated: `r format(Sys.time(), '%d, %B, %Y at %H:%M')`"
output:
    html_document:
        toc: true
        toc_float: true
        code_folding: hide
        theme: cerulean
params:
    countFilePath_SALMON: "countFilePath_SALMON"
    output: "output"
    coldata: "coldata"
    annotated: "annotated"
    formula: "formula"
    min_count: 10
    fdr: 0.05
    conditionN: 2
    log2FCT: 2
    topNGenes: 20
    t2gPath: "t2gPath"
---

# Setup

## Intruduction

This report is designed to be used with the SALMON pipeline.
It will take the output of the SALMON pipeline and perform differential gene expression analysis using DESeq2 and edgeR.
The report will also generate a number of plots to help visualise the data.

## Samplesheet and count data content

Check for samplesheet and count data content.

```{r setup, include=FALSE}
knitr::opts_knit$set(
  root.dir = "../"
  )
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
  )
```

```{r setup_libPrep, include=TRUE}
suppressPackageStartupMessages({
  library(DESeq2)
  library(DT)
  library(edgeR)
  library(EnhancedVolcano)
  library(ggplot2)
  library(ggvenn)
  library(gplots)
  library(grDevices)
  library(gridExtra)
  library(plotly)
  library(RColorBrewer)
  library(reshape2)
  library(tximport)
  library(tximportData)

  library(statmod)
  library(apeglm)
})

# Print R version
print(paste("R version:", R.version.string))


options(
  DT.options = list(
    columnDefs = list(list(className = 'dt-center', targets = "_all")),
    scrollX = TRUE,
    # pageLength = 5,
    language = list(search = 'Filter:')
    )
  )


# Numerical values
fdr <- params$fdr
min_count <- params$min_count
conditionN <- params$conditionN
topNGenes <- params$topNGenes
log2FCT <- params$log2FCT

# Input data
countFilePath_SALMON <- params$countFilePath_SALMON
coldata <- params$coldata

t2gPath <- params$t2gPath
annotated <- params$annotated
# annotated$ensembl_gene_id
# Kepp only unique genes, or the first gene if there are multiple
annotated <- annotated[!duplicated(annotated$ensembl_gene_id), ]

formula <- params$formula

r_output <- params$output
if (r_output == "") {
    r_output <- "."
}

salmon_folder <- "/SALMON"
# Create SALMON folder if it does not exist
if (!dir.exists(paste0(r_output, salmon_folder))) {
    dir.create(paste0(r_output, salmon_folder))
}
# Output paths
SALMON_edgeR.csv <- paste0(r_output, salmon_folder, "/SALMON_edgeR.csv")
SALMON_edgeR_counts.csv <- paste0(r_output, salmon_folder, "/SALMON_edgeR_counts.csv")
SALMON_DESeq2.csv <- paste0(r_output, salmon_folder, "/SALMON_DESeq2.csv")
SALMON_DESeq2_counts.csv <- paste0(r_output, salmon_folder, "/SALMON_DESeq2_counts.csv")

common_df.csv <- paste0(r_output, salmon_folder, "/common_sig_genes.csv")
common_df_both.csv <- paste0(r_output, salmon_folder, "/common_sig_DE_genes.csv")

# Data inport
salmon_data <- paste0(countFilePath_SALMON, "/", coldata[[1]], "/quant.sf")
names(salmon_data) <- coldata[[1]]
# Check if all samples exists
exists <- sapply(salmon_data, file.exists)
# Remove non-existing samples
salmon_data <- salmon_data[exists]

# Find all "/quant.sf" files
salmon_data_present <- list.files(countFilePath_SALMON, recursive = TRUE, pattern = "quant.sf", full.names = TRUE)

tx2gene <- read.delim(t2gPath, header = FALSE)
tx2gene <- tx2gene[c(1,2)]

countdata_SALMON <- tximport(salmon_data, type = "salmon", tx2gene = tx2gene)
countdata_SALMON <- as.data.frame(countdata_SALMON$counts)

colnames(countdata_SALMON) <- gsub(colnames(countdata_SALMON), pattern = "\\.", replacement = "-")
rownames(countdata_SALMON) <- sub('\\.[0-9]*$', '', rownames(countdata_SALMON))

datatable(
    coldata,
    caption = "Samplesheet",
    # rownames = NA,
    extensions = 'FixedColumns'
)
# Check if all samples are present in the countdata
# cat(paste("All samples:", paste(coldata[, 1], collapse = "\n"), sep = "\n"))
cat(paste("Countdata samples:", paste(colnames(countdata_SALMON), collapse = "\n"), sep = "\n"))
cat(paste("Extra samples in samplesheet:", 
    paste(names(exists)[!exists], collapse = "\n"), sep = "\n"))
cat(paste("Extra samples in countdata:",
    paste(salmon_data_present[!salmon_data_present %in% salmon_data], collapse = "\n"), sep = "\n"))

# Remove samples that are not present in the countdata but are in countdata
coldata <- coldata[salmon_data %in% salmon_data_present, ]
# Remove samples that are not present in the samplesheet but are in countdata
countdata_SALMON <- countdata_SALMON[, coldata[, 1]]

test_samples <- data.frame(
    Samplesheet_samples = coldata[, 1],
    Countdata_samples = colnames(countdata_SALMON)
)
datatable(
    test_samples,
    caption = "Samples that are present in both samplesheet and countdata",
    # rownames = NA,
    extensions = 'FixedColumns'
)

# Color & size
col_vector <- grDevices::colors()[grep('gr(a|e)y', grDevices::colors(), invert = T)]
n <- length(unique(coldata[, 1])) + length(unique(coldata[, conditionN]))
sub_col_vector <- sample(size = n, col_vector)

```

The formula used to model gene expression in this analysis is `r paste(formula)`.

# SALMON report

## Raw data exploration

Gene with the highest expression across all samples

```{r SALMON_topGene}
topGene <- which.max(rowSums(countdata_SALMON))
print(topGene)

datatable(
    t(countdata_SALMON[topGene, ]),
    # rownames = NA,
    extensions = 'FixedColumns'
)
```

We can investigate this data a bit more using some of the basic R functions before going on to use more sophisticated analysis tools.
First we will calculate the mean for each gene for each condition and plot them comparativly.

```{r SALMON_get_means}
# Get unique values in the "Treatment" column
unique_treatments <- unique(coldata[, conditionN])

gene_means <- sapply(unique_treatments, function(treatment) {
    rowMeans(countdata_SALMON[, coldata[, 1][coldata[, conditionN] == treatment]])
})
gene_means <- data.frame(gene_means)

# Generate all pairwise combinations of mean names
mean_combinations <- combn(unique_treatments, 2, simplify = FALSE)

plot_list <- list()

# Generate plots for each mean combination
for (i in seq_along(mean_combinations)) {
    mean_x <- mean_combinations[[i]][1]
    mean_y <- mean_combinations[[i]][2]

    # Create a scatterplot for the mean comparison
    p <- ggplot(gene_means, aes_string(x = mean_x, y = mean_y)) +
        geom_point() +
        scale_y_continuous(trans = 'log10') +
        scale_x_continuous(trans = 'log10') +
        labs(x = mean_x, y = mean_y) +
        theme_bw()

    # Store the plot in the list
    plot_list[[i]] <- p
}

# Arrange and display the plots
grid.arrange(grobs = plot_list, nrow = length(mean_combinations))
```

### Raw differential gene expression

We can find candidate differentially expressed genes by looking for genes with a large change between samples. 
A common threshold used is log2 fold change more than `r log2FCT` or less than `r -log2FCT`. 
We will calculate log2 fold change for all the genes and colour the genes with log2 fold change of more than 2 or 
less than -2 on the plot.

First, check for genes with a mean expression of 0. Putting zeroes into the log2 fold change calculation will produce NAs, 
so we might want to remove these genes. Note: this is for mathematical reasons, although different software may produce 
different results when you try to do `log2(0)`.

how many genes have a mean 0?
```{r SALMON_count_expressed_0}
print(
    apply(gene_means, 2, function(x) {
        sum(x <= 0)
    })
)
```

how many genes have a mean above 0?

```{r SALMON_count_expressed_1}
print(
    apply(gene_means, 2, function(x) {
        sum(x > 0)
    })
)
```

```{r SALMON_subset_nonzero}
print(paste("Number of genes befor filtering:", nrow(gene_means)))

# Remove genes with a mean of 0
gene_means[gene_means == 0] <- NA
gene_means <- na.omit(gene_means)

print(paste("Number of genes after filtering:", nrow(gene_means)))
```

Mathematically things work out better for us when we test things on the log scale.
On the absolute scale, upregulation goes from 1 to infinity, while downregulation is bounded by 0 and 1. On the log scale,
upregulation goes from 0 to infinity, and downregulation goes from 0 to negative infinity. Let's compute a log-base-2 of the fold change.

When we do this we'll see some `Inf` and `-Inf` values. This is what happens when we take `log2(Inf)` or `log2(0)`.

```{r SALMON_log2FC}
gene_log <- data.frame(matrix(nrow = nrow(gene_means), ncol = 0))
plot_list_add <- list()

for (i in seq_along(mean_combinations)) {
    mean_x <- mean_combinations[[i]][1]
    mean_y <- mean_combinations[[i]][2]

    comb_name <- paste0("log2FC_", mean_x, "_", mean_y)
    gene_log[, comb_name] <- log2(gene_means[, mean_x] / gene_means[, mean_y])

    # Report
    print(paste0(comb_name, " - Differently expresed: ", sum(abs(gene_log) > 2)))
    print(paste0(comb_name, " - Underexpresed expresed: ", sum(gene_log < -2)))
    print(paste0(comb_name, " - Overexpresed expresed: ", sum(gene_log > 2)))

    gene_sig <- gene_means[abs(gene_log) > log2FCT, ]

    # color significant genes
    plot_list[[i]] <- plot_list[[i]] +
        geom_point(data = gene_sig, aes(col = "red"), size = 2) +
        theme(legend.position = "none")
}

# Arrange and display the plots
grid.arrange(grobs = plot_list, nrow = length(mean_combinations))
```

```{r SALMON_log2FC_top}
gene_means <- cbind(gene_means, gene_log)
gene_means$ID <- rownames(gene_means)
gene_means <- merge(gene_means, annotated, by.x = "ID", by.y = "ensembl_gene_id")

indx <- order(abs(gene_log[, 1]), decreasing = TRUE)
gene_means <- gene_means[indx, ]

# print(head(gene_means, topNGenes))
datatable(
    gene_means,
    rownames = NA,
    extensions = 'FixedColumns'
)
# TODO: save?
```


## edgeR analysis

### edgeR DEList


For easy manipulation, we put the data into a DGEList object

```{r SALMON_edgeR_DEList}
# Create DEList
y_SALMON <- DGEList(counts = countdata_SALMON)
# print(y_SALMON)

all_genes_unique <- length(rownames(y_SALMON$counts)) == length(unique(rownames(y_SALMON$counts)))
print(paste0("Are all genes unique in counts: ", all_genes_unique))
```

### edgeR filter

All datasets will include a mix of genes that are expressed and those that are not expressed. Whilst it is of interest 
to examine genes that are expressed in one condition but not in another, some genes are unexpressed throughout all samples.
Filter genes with at least `r min_count` counts.

```{r SALMON_edgeR_Filtration}
cpm <- cpm(y_SALMON)
lcpm <- cpm(y_SALMON, log = TRUE)
L <- mean(y_SALMON$samples$lib.size) * 1e-6
M <- median(y_SALMON$samples$lib.size) * 1e-6
# summary(lcpm)
print(paste("mean:", L, " --- median:", M))

min_count <- 10
keep_SALMON <- filterByExpr(y_SALMON, min.count = min_count)
missing_gene <- table(rowSums(y_SALMON$counts == 0))
missing_sample_0 <- colSums(y_SALMON$counts == 0)
missing_sample_min <- colSums(y_SALMON$counts < min_count)

print(paste("Max:", length(keep_SALMON)))
print(paste("To keep:", sum(keep_SALMON)))
print(paste("To keep [%]:", round((sum(keep_SALMON) / length(keep_SALMON)) * 100), "%"))

# invisible(dev.off())
# par(mfrow = c(1, 1))
plot(missing_gene)
# invisible(dev.off())

y_SALMON <- y_SALMON[keep_SALMON, , keep.lib.sizes = FALSE]
```

Genes that do not have a worthwhile number of reads in any sample should be filtered out of the downstream analyses. 
There are several reasons for this. From a biological point of view, genes that not expressed at a biologically meaningful 
level in any condition are not of interest and are therefore best ignored. From a statistical point of view, removing low count genes 
allows the mean-variance relationship in the data to be estimated with greater reliability and also reduces the number of 
statistical tests that need to be carried out in downstream analyses looking at differential expression.

```{r SALMON_edgeR_Missing, fig.width=10, fig.height=5}
missing_sample_0_df <- data.frame(
  sample = names(missing_sample_0),
  genes = missing_sample_0
)
missing_sample_min_df <- data.frame(
  sample = names(missing_sample_min),
  genes = missing_sample_min
)

# Missing plot
p1 <- ggplot(data = missing_sample_0_df, aes(x = sample, y = genes)) +
  geom_bar(stat = "identity") +
  geom_hline(aes(yintercept = mean(genes), col = "mean")) +
  geom_hline(aes(yintercept = median(genes), col = "median")) +
  theme_bw()
ggplotly(p1)
invisible(dev.off())

p2 <- ggplot(data = missing_sample_min_df, aes(x = sample, y = genes)) +
  geom_bar(stat = "identity") +
  geom_hline(aes(yintercept = mean(genes), col = "mean")) +
  geom_hline(aes(yintercept = median(genes), col = "median")) +
  theme_bw()
ggplotly(p2)
# invisible(dev.off())
```

Plotting the distribution log-CPM values shows that a sizeable proportion of genes within each sample are either 
unexpressed or lowly-expressed with log-CPM values that are small or negative.

```{r SALMON_edgeR_Missing_diss, fig.width=10, fig.height=5}
samplenames <- coldata[[1]]
lcpm.cutoff <- log2(10 / M + 2 / L)
nsamples <- ncol(y_SALMON)
col <- brewer.pal(nsamples, "Paired")
par(mfrow = c(1, 2))
plot(density(lcpm[, 1]), col = col[1], lwd = 2, ylim = c(0, 0.26), las = 2, main = "", xlab = "")
title(main = "A. Raw data", xlab = "Log-cpm")
abline(v = lcpm.cutoff, lty = 3)
for (i in 2:nsamples) {
  den <- density(lcpm[, i])
  lines(den$x, den$y, col = col[i], lwd = 2)
}
# legend("topright", samplenames, text.col = col, bty = "n")

lcpm <- cpm(y_SALMON, log = TRUE)
plot(density(lcpm[, 1]), col = col[1], lwd = 2, ylim = c(0, 0.26), las = 2, main = "", xlab = "")
title(main = "B. Filtered data", xlab = "Log-cpm")
abline(v = lcpm.cutoff, lty = 3)
for (i in 2:nsamples) {
  den <- density(lcpm[, i])
  lines(den$x, den$y, col = col[i], lwd = 2)
}
invisible(dev.off())
# legend("topright", samplenames, text.col = col, bty = "n")
```

### edgeR normalization

TMM normalization is applied to this dataset to account for compositional difference between
the libraries. For example, samples processed in the first batch of an experiment can have higher 
expression overall when compared to samples processed in a second batch.

```{r SALMON_edgeR_normalization}
y_SALMON <- calcNormFactors(y_SALMON, design, method = "TMM")

x2 <- y_SALMON
x2$samples$norm.factors <- 1
x2$counts[, 1] <- ceiling(x2$counts[, 1] * fdr)
x2$counts[, 2] <- x2$counts[, 2] * 5

par(mfrow = c(1, 2))
lcpm <- cpm(x2, log = TRUE)
boxplot(lcpm, las = 2, col = col, main = "")
title(main = "Unnormalised data", ylab = "Log-cpm")
x2 <- calcNormFactors(x2)

lcpm <- cpm(x2, log = TRUE)
boxplot(lcpm, las = 2, col = col, main = "")
title(main = "Normalised data", ylab = "Log-cpm")

invisible(dev.off())
```


```{r SALMON_edgeR_normalization_plot_silent, fig.show='hide'}
par(mfrow = c(1, 1))
mds <- plotMDS(
  y_SALMON,
  labels = coldata[[1]],
  col = as.numeric(as.factor(coldata[[1]])),
  cex = 0.75
)
```
```{r SALMON_edgeR_normalization_plot}
toplot <- data.frame(
    Dim1 = mds$x,
    Dim2 = mds$y,
    Group = as.factor(coldata[[conditionN]]),
    Sample = coldata[['Patient_ID']]
    )

gp <- ggplot(toplot, aes(Dim1, Dim2, color = Group))
gp <- gp +
    geom_hline(aes(yintercept = 0), colour = "grey") +
    geom_vline(aes(xintercept = 0), colour = "grey") +
    geom_point(size = 4) +
    xlab("Dimension 1") +
    ylab("Dimension 2") +
    ggtitle("MDS plot\n") +
    theme_bw()

gp <- gp +
geom_point(aes(toplot$Dim1, toplot$Dim2, color = toplot$Sample), size = 2) +
scale_color_manual(values = sub_col_vector)

ggplotly(gp)
```


The first step of an analysis should be to examine the samples for outliers and for other
relationships. The function plotMDS produces a plot in which distances between samples
correspond to leading biological coefficient of variation (BCV) between those samples.


### edgeR estimate dispersion

The square root of the common dispersion gives the coefficient of variation of biological
variation.
The dispersion estimates can be viewed in a BCV plot.

```{r SALMON_edgeR_dispersion}
y_SALMON <- estimateDisp(y_SALMON, design, robust = TRUE)
y_SALMON$common.dispersion

plotBCV(y_SALMON)
```

Here the common dispersion is found to be `r round(y_SALMON$common.dispersion, digits = 5)`, so the coefficient of biological
variation is around `r round(sqrt(y_SALMON$common.dispersion), digits = 5)`.


### edgeR fitting & testing

```{r SALMON_edgeR_Test1}
# Fit genewise glms
fit_SALMON <- glmQLFit(y_SALMON, design)

# Conduct likelihood ratio tests for treated vs pre-tretment tissue differences
lrt <- glmLRT(fit_SALMON)

plotMD(lrt)
abline(h = c(-1, 1), col = "blue")
```


### edgeR correction

The genewise tests are for treated vs pre-treated differential expression, adjusting for baseline
differences between the patients. The tests can be viewed as analogous to paired
t-tests. The top DE tags have tiny p-values and FDR values, as well as large fold changes.

```{r SALMON_edgeR_correction}
results_SALMON_edgeR <- topTags(lrt, adjust.method = "BH", n = nrow(lrt$table))

results_SALMON_edgeR <- na.omit(results_SALMON_edgeR)
results_SALMON_edgeR <- data.frame(results_SALMON_edgeR)
results_SALMON_edgeR$ID <- rownames(results_SALMON_edgeR)

head(results_SALMON_edgeR[, -ncol(results_SALMON_edgeR)])
```

The total number of differentially expressed genes at `r fdr` FDR:

```{r SALMON_edgeR_correction_stats}
summary(decideTests(lrt))
```

### edgeR P-values distribution

Histogram of the unadjusted and adjusted p-values.

```{r SALMON_edgeR_hist}
hp <- ggplot(results_SALMON_edgeR) +
  geom_histogram(aes(PValue, fill = "p-value"), colour = "grey20", alpha = 0.5, stat = "bin") +
  geom_histogram(aes(FDR, fill = "padj"), colour = "grey20", alpha = 0.5, stat = "bin") +
  scale_fill_manual(
    name = "group", values = c("p-value" = "steelblue", "padj" = "grey20"), labels = c("a" = "p-value", "b" = "padj")
    ) +
  geom_vline(xintercept = fdr, colour = "red") +
  theme_bw()

ggplotly(hp)
```

```{r SALMON_edgeR_hist_table}
## Split features by different adjusted p-value cutoffs
padj_table <- lapply(c(1e-04, 0.001, 0.01, 0.025, 0.05, 0.1, 0.2, 0.3, 0.4, 0.5,
    0.6, 0.7, 0.8, 0.9, 1), function(x) {
    data.frame('Cut' = x, 'Count' = sum(results_SALMON_edgeR$FDR <= x, na.rm = TRUE))
})
padj_table <- do.call(rbind, padj_table)

datatable(
  padj_table,
  rownames = NA,
  extensions = 'FixedColumns'
)
```

```{r SALMON_edgeR_results_summary}
print(paste(
  "Number of genes with significant p-value:",
  sum(results_SALMON_edgeR$PValue < fdr)
  ))

EDGER_Nsig <- sum(results_SALMON_edgeR$FDR < fdr)
print(paste(
  "Number of genes with significant adjusted p-value:",
  EDGER_Nsig
  ))
```

Annotate discovered genes.
```{r SALMON_edgeR_annotate}
# define biomart object
results_SALMON_edgeR <- merge(results_SALMON_edgeR, annotated, by.x = "ID", by.y = "ensembl_gene_id")
results_SALMON_edgeR <- results_SALMON_edgeR[order(results_SALMON_edgeR$FDR), ]
results_SALMON_edgeR <- results_SALMON_edgeR[results_SALMON_edgeR$PValue <= fdr, ]

write.table(results_SALMON_edgeR, file = SALMON_edgeR.csv, append = FALSE, quote = FALSE, sep = "\t", row.names = FALSE)

counts <- data.frame(cpm(lrt))
counts$ID <- rownames(counts)
results_SALMON_edgeR_counts <- merge(results_SALMON_edgeR, counts, by = "ID")
write.table(results_SALMON_edgeR_counts,
  file = SALMON_edgeR_counts.csv,
  append = FALSE, quote = FALSE, sep = "\t", row.names = FALSE
)
```

### edgeR volcano plot

[Volcano plots](https://en.wikipedia.org/wiki/Volcano_plot_(statistics)) are another (sometimes preferred) way of visulizing the DE results. 
It clearly identifies the highest changing genes from the results.

```{r SALMON_edgeR_VC}
EnhancedVolcano(
    results_SALMON_edgeR,
    lab = "",
    x = 'logFC',
    y = 'FDR',
    title = 'SALMON-EdgeR results',
    subtitle = "",
    legendPosition = "top",
    pCutoff = fdr
    # title = 'SALMON-DESeq results'
    # subtitle = "Differential expression"
)
```

### edgeR results

```{r SALMON_edgeR_results}
datatable(
    results_SALMON_edgeR,
    rownames = NA,
    extensions = 'FixedColumns'
)
```


## DESeq2 analysis

DESeq2 is an R package for analyzing count-based NGS data like RNA-seq. It is available 
from [Bioconductor](http://www.bioconductor.org/). Bioconductor is a project to provide tools for analysing 
high-throughput genomic data including RNA-seq, ChIP-seq and arrays. 
You can explore Bioconductor packages [here](http://www.bioconductor.org/packages/release/BiocViews.html#___Software).
Analysis sourced from 
[Vignette](http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#input-data) 
and [Github](https://gist.github.com/stephenturner/f60c1934405c127f09a6).

### Preparing outputs for exploration

Create a coldata frame and instantiate the DESeqDataSet

```{r SALMON_DESeq_DESeqDataSet}
dds <- DESeqDataSetFromMatrix(countData = round(countdata_SALMON), colData = coldata, design = design)

dds
```

### DESeq filter

```{r SALMON_DESeq_filter}
keep <- rowSums(counts(dds)) >= min_count
dds <- dds[keep, ]
dds
```


### DESeq pipeline

```{r SALMON_DESeq_Pipe}
dds <- DESeq(dds)
dds
```

### DESeq dispersion

This plot shows the fitting of dispersion estimates for all genes by DESeq2. A better fit would mean 
more reliable results.

```{r SALMON_DESeq_dispersion}
plotDispEsts(dds, main = "Dispersion plot")
```

### DESeq regularization

Regularized log transformation for clustering/heatmaps, etc

```{r SALMON_DESeq_regularization}
rld <- rlogTransformation(dds)
# head(assay(rld))
hist(assay(rld))
```

### DESeq Heatmap

This plot shows how samples are clustered based on their euclidean distance using the regularized 
log transformed count data. This figure gives an overview of how the samples are hierarchically clustered. 
It is a complementary figure to the PCA plot.

```{r SALMON_DESeq_hm}
condition <- factor(coldata[[conditionN]])
mycols <- brewer.pal(8, "Dark2")[1:length(unique(condition))]

# Sample distance heatmap
sampleDists <- as.matrix(dist(t(assay(rld))))
heatmap.2(as.matrix(sampleDists),
  key = F, trace = "none",
  col = colorpanel(100, "#1109ee", "#ffffff"),
  ColSideColors = mycols[condition], RowSideColors = mycols[condition],
  margin = c(10, 10), main = "Sample Distance Matrix"
)
```

### DESeq PCA

PCA plot using the first two principal components that explain the variability in the data using the regularized 
log count data. If you are unfamiliar with principal component analysis, 
you might want to check this [interactive explanation](http://setosa.io/ev/principal-component-analysis/). 

```{r SALMON_DESeq_pca}
# n <- length(unique(coldata[, 1])) + length(unique(coldata[, conditionN]))
PCAdata <- DESeq2::plotPCA(rld, intgroup = colnames(coldata)[conditionN], returnData = TRUE)
percentVar <- round(100 * attr(PCAdata, "percentVar"))

mat <- match(PCAdata$name, coldata[, 1])
PCAdata$sample <- as.character(coldata$Patient_ID[mat])

gp <- ggplot(PCAdata, aes(PC1, PC2, color = condition))
gp <- gp +
    geom_hline(aes(yintercept = 0), colour = "grey") +
    geom_vline(aes(xintercept = 0), colour = "grey") +
    geom_point(size = 4) +
    xlab(paste0("PC1: ", percentVar[1], "% variance")) +
    ylab(paste0("PC2: ", percentVar[2], "% variance")) +
    ggtitle("PCA\n")+
    theme_bw()

gp <- gp +
geom_point(aes(PCAdata$PC1, PCAdata$PC2, color = PCAdata$sample), size = 2) +
scale_color_manual(values = sub_col_vector)

ggplotly(gp)
```

In this case, the first and second principal component 
explain `r percentVar[1]` and `r percentVar[2]` percent of the variance respectively.

### DESeq DE genes

Get differential expression results

```{r SALMON_DESeq_genes}
res <- results(dds)
res <- na.omit(res)
```

How many genes are significant?

```{r SALMON_DESeq_siggenes}
table(res$padj < fdr)
```

### DESeq MA plot

MA plots compare the mean of the normalized counts against the log fold change. 
They show one point per feature. 
The points are shown in blue if the feature has an adjusted p-value less than alpha, 
that is, the statistically significant features are shown in blue.

```{r SALMON_DESeq_MA}
maplot <- function(res, thresh = fdr, labelsig = FALSE, textcx = 1, ...) {
  with(res, plot(
    baseMean, log2FoldChange, pch = 20, cex = .8, log = "x",
    xlab = "Mean of Normalized Counts", ylab = "Log2 Fold Change", ...
    ))
  with(subset(res, padj < thresh), points(baseMean, log2FoldChange, col = "blue", pch = 20, cex = 1))
  if (labelsig) {
    require(calibrate)
    with(subset(res, padj < thresh), textxy(baseMean, log2FoldChange, labs = Gene, cex = textcx, col = 2))
  }
}

maplot(res, main = "MA Plot")
# invisible(dev.off())
# TODO: Plot not showing

# Package
plotMA(
  res, ylim = c(-2, 2), cex = .8,
  xlab = "Mean of Normalized Counts",
  ylab = "Log Fold Change"
  )
invisible(dev.off())

# Shrinken low values
resAshT <- lfcShrink(dds, coef = colnames(design)[ncol(design)], type = "apeglm")
plotMA(
  resAshT, ylim = c(-3, 3), cex = .8,
  xlab = "Mean of Normalized Counts",
  ylab = "Shrinken Log Fold Change"
  )
abline(h = c(-1, 1), col = "dodgerblue", lwd = 2)
invisible(dev.off())
```

```{r SALMON_DESeq_formating}
## Order by adjusted p-value
res <- res[order(res$padj), ]
results_SALMON_DESeq2 <- as.data.frame(res)
results_SALMON_DESeq2$ID <- rownames(results_SALMON_DESeq2)
## Merge with normalized count data
# results_SALMON_DESeq2 <- merge(as.data.frame(res), as.data.frame(counts(dds, normalized = TRUE)), by = "row.names", sort = FALSE)
# names(results_SALMON_DESeq2)[1] <- "ID"
```

### DESeq P-values distribution

Histogram of the unadjusted and adjusted p-values.

```{r SALMON_DESeq_hist}
hp <- ggplot(results_SALMON_DESeq2) +
  geom_histogram(aes(pvalue, fill = "p-value"), colour = "grey20", alpha = 0.5, stat = "bin") +
  geom_histogram(aes(padj, fill = "padj"), colour = "grey20", alpha = 0.5, stat = "bin") +
  scale_fill_manual(
    name = "group", values = c("p-value" = "steelblue", "padj" = "grey20"), labels = c("a" = "p-value", "b" = "padj")
    ) +
  geom_vline(xintercept = fdr, colour = "red") +
  theme_bw()

ggplotly(hp)
```

```{r SALMON_DESeq_hist_table}
## Split features by different adjusted p-value cutoffs
padj_table <- lapply(c(1e-04, 0.001, 0.01, 0.025, 0.05, 0.1, 0.2, 0.3, 0.4, 0.5,
    0.6, 0.7, 0.8, 0.9, 1), function(x) {
    data.frame('Cut' = x, 'Count' = sum(results_SALMON_DESeq2$padj <= x, na.rm = TRUE))
})
padj_table <- do.call(rbind, padj_table)

datatable(
  padj_table,
  rownames = NA,
  extensions = 'FixedColumns'
)
```

```{r SALMON_DESeq_results}
print(paste(
  "Number of genes with significant p-value:",
  sum(results_SALMON_DESeq2$pvalue < fdr)
  ))

DESEQ2_Nsig <- sum(results_SALMON_DESeq2$padj < fdr)
print(paste(
  "Number of genes with significant adjusted p-value:",
  DESEQ2_Nsig
  ))
```

```{r SALMON_DESeq_results_annotation}
# Annotate gnes and save to table
results_SALMON_DESeq2 <- merge(results_SALMON_DESeq2, annotated, by.x = "ID", by.y = "ensembl_gene_id")
results_SALMON_DESeq2 <- results_SALMON_DESeq2[order(results_SALMON_DESeq2$padj), ]
write.table(results_SALMON_DESeq2, file = SALMON_DESeq2.csv, append = FALSE, quote = FALSE, sep = "\t", row.names = FALSE)

counts <- data.frame(counts(dds, normalized = TRUE))
counts$ID <- rownames(counts)
results_SALMON_DESeq_counts <- merge(results_SALMON_DESeq2, counts, by = "ID")
write.table(results_SALMON_DESeq_counts, file = SALMON_DESeq2_counts.csv,
  append = FALSE, quote = FALSE, sep = "\t", row.names = FALSE
  )
```

### DESeq volcano plot

[Volcano plots](https://en.wikipedia.org/wiki/Volcano_plot_(statistics)) are another (sometimes preferred) way of visulizing the DE results. 
It clearly identifies the highest changing genes from the results.

```{r SALMON_DESeq_VC}
EnhancedVolcano(
    results_SALMON_DESeq2,
    lab = "",
    x = 'log2FoldChange',
    y = 'padj',
    title = 'SALMON-DESeq results',
    subtitle = "",
    legendPosition = "top",
    pCutoff = fdr
    # title = 'SALMON-DESeq results'
    # subtitle = "Differential expression"
)
```

### DESeq results

```{r SALMON_DESeq_DT}
datatable(
    results_SALMON_DESeq2,
    rownames = NA,
    extensions = 'FixedColumns'
)
```

# Combined

Venn Diagram of the DE genes from the different methods.
For further analysis, we will use only the genes that are significant in the DESeq2 and edgeR methods.

Comparison of DE genes acording to the adjusted p-value cutoff (`r fdr`).

```{r Combined_prep}
genes_DESeq2 <- results_SALMON_DESeq2$ID[results_SALMON_DESeq2$padj > fdr]
genes_edgeR <- results_SALMON_edgeR$ID[results_SALMON_edgeR$FDR > fdr]
genes_simple <- gene_means$ID[gene_means[, which(grepl("log2FoldChange", colnames(results_SALMON_DESeq2)))[1]] > log2FCT]

D <- list(
  genes_DESeq2 = genes_DESeq2,
  genes_edgeR = genes_edgeR,
  genes_simple = genes_simple
  )

ggvenn(
  D,
  show_percentage = TRUE,
  fill_color = sub_col_vector
  )

# Create and save a dataframe with the common genes
genes_common <- genes_DESeq2[genes_DESeq2 %in% genes_edgeR]

# results_SALMON_DESeq2[results_SALMON_DESeq2$ID %in% "ENSG00000230417", ]

row.names(results_SALMON_DESeq2) <- results_SALMON_DESeq2$ID
row.names(results_SALMON_edgeR) <- results_SALMON_edgeR$ID

results_SALMON_DESeq2_common <- results_SALMON_DESeq2[genes_common, ]
results_SALMON_edgeR_common <- results_SALMON_edgeR[genes_common, ]

common_df <- data.frame(
  ID = row.names(results_SALMON_DESeq2_common),
  log2FoldChange_DESeq2 = results_SALMON_DESeq2_common$log2FoldChange,
  log2FoldChange_edgeR = results_SALMON_edgeR_common$logFC,
  pvalue_DESeq2 = results_SALMON_DESeq2_common$pvalue,
  pvalue_edgeR = results_SALMON_edgeR_common$PValue,
  padj_DESeq2 = results_SALMON_DESeq2_common$padj,
  padj_edgeR = results_SALMON_edgeR_common$FDR,
  hgnc_symbol = results_SALMON_edgeR_common$hgnc_symbol,
  description = results_SALMON_edgeR_common$description
  )

datatable(
    common_df,
    rownames = NA,
    extensions = 'FixedColumns'
)

write.table(common_df, file = common_df.csv, append = FALSE, quote = FALSE, sep = "\t", row.names = FALSE)
```

Comparison of DE genes acording to the adjusted p-value cutoff (`r fdr`) and log2FoldChange (`r log2FCT`).

```{r Combined_both} 
genes_DESeq2 <- results_SALMON_DESeq2$ID[results_SALMON_DESeq2$padj > fdr & abs(results_SALMON_DESeq2$log2FoldChange) > log2FCT]
genes_edgeR <- results_SALMON_edgeR$ID[results_SALMON_edgeR$FDR > fdr & abs(results_SALMON_edgeR$logCPM) > log2FCT]
genes_simple <- gene_means$ID[gene_means[, which(grepl("log2FoldChange", colnames(results_SALMON_DESeq2)))[1]] > log2FCT]

D <- list(
  genes_DESeq2 = genes_DESeq2,
  genes_edgeR = genes_edgeR,
  genes_simple = genes_simple
  )

ggvenn(
  D,
  show_percentage = TRUE,
  fill_color = sub_col_vector
  )

# Create and save a dataframe with the common genes
genes_common <- genes_DESeq2[genes_DESeq2 %in% genes_edgeR]

row.names(results_SALMON_DESeq2) <- results_SALMON_DESeq2$ID
row.names(results_SALMON_edgeR) <- results_SALMON_edgeR$ID

results_SALMON_DESeq2_common <- results_SALMON_DESeq2[genes_common, ]
results_SALMON_edgeR_common <- results_SALMON_edgeR[genes_common, ]

common_df <- data.frame(
  ID = row.names(results_SALMON_DESeq2_common),
  log2FoldChange_DESeq2 = results_SALMON_DESeq2_common$log2FoldChange,
  log2FoldChange_edgeR = results_SALMON_edgeR_common$logFC,
  pvalue_DESeq2 = results_SALMON_DESeq2_common$pvalue,
  pvalue_edgeR = results_SALMON_edgeR_common$PValue,
  padj_DESeq2 = results_SALMON_DESeq2_common$padj,
  padj_edgeR = results_SALMON_edgeR_common$FDR,
  hgnc_symbol = results_SALMON_edgeR_common$hgnc_symbol,
  description = results_SALMON_edgeR_common$description
  )

datatable(
    common_df,
    rownames = NA,
    extensions = 'FixedColumns'
)

write.table(common_df, file = common_df_both.csv, append = FALSE, quote = FALSE, sep = "\t", row.names = FALSE)
```


# Sources

https://github.com/gantunes00/rnaseq-data-analysis-brb/blob/master/02-differential-expression.md


https://4va.github.io/biodatasci/r-rnaseq-airway.html


https://www.bioconductor.org/packages/release/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf


http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html


https://github.com/maxplanck-ie/snakepipes


https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4937821/
